# Provincial Yield table - NSYT and MSYT combined table
## Overview

This script generates a combined yield table containing:

- Conifer merchantable volume (`MVcon_*`)
- Deciduous merchantable volume (`MVdec_*`)
- Conifer dominant height (`HTcon_*`)
- Deciduous dominant height (`HTdec_*`)

The workflow combines:

1. VDYP-based yield curves (NSYT workflow)
2. MSYT yield curves

The final output is exported as a single yield table for downstream modeling.

```{r setup, include=FALSE}
#globally to the entire document
knitr::opts_chunk$set(
  eval = FALSE,   
  echo = TRUE    
)
```

```{r libraries}
## Load Libraries
library(data.table)
library(dplyr)
library(tidyr)
```

## Define Deciduous Species
Species in this list are treated as deciduous.
```{r deciduous-species}
decid <- c(
  "AC","ACB","ACT","AT","AD","AX",
  "DR","E","EA","EB","EE","EP","ES","EW",'LS',
  "EXP","XH","G","GP","KC","M","MB","MN","MV","Q","QG","XH",
  "WT","WS","WP","WD","WB","WA","W", 
  "VP","VB","V","U","RA"
)
```

`NoDecidVol_calculator()` calculates conifer-only projected volume by removing deciduous species from VDYP projected species volumes.
```{r function}
NoDecidVol_calculator <- function (sp_list,
                                   vol_list,
                                   decidsp,
                                   multilayer = FALSE) {

  # Standardize species slot names
  sp_list[spcode %in% "SPECIES_1_CODE", `:=`(spcode, "sp1")]
  sp_list[spcode %in% "SPECIES_2_CODE", `:=`(spcode, "sp2")]
  sp_list[spcode %in% "SPECIES_3_CODE", `:=`(spcode, "sp3")]
  sp_list[spcode %in% "SPECIES_4_CODE", `:=`(spcode, "sp4")]
  sp_list[spcode %in% "SPECIES_5_CODE", `:=`(spcode, "sp5")]
  sp_list[spcode %in% "SPECIES_6_CODE", `:=`(spcode, "sp6")]

  # Remove deciduous species
  sp_list2 <- sp_list[
    !is.na(sp) &
      sp != "" &
      !sp %in% decidsp
  ]

  # Replace NA projected volumes with 0
  vol_list[is.na(PRJ_SP1_VOL_DWB), `:=`(PRJ_SP1_VOL_DWB, 0)]
  vol_list[is.na(PRJ_SP2_VOL_DWB), `:=`(PRJ_SP2_VOL_DWB, 0)]
  vol_list[is.na(PRJ_SP3_VOL_DWB), `:=`(PRJ_SP3_VOL_DWB, 0)]
  vol_list[is.na(PRJ_SP4_VOL_DWB), `:=`(PRJ_SP4_VOL_DWB, 0)]
  vol_list[is.na(PRJ_SP5_VOL_DWB), `:=`(PRJ_SP5_VOL_DWB, 0)]
  vol_list[is.na(PRJ_SP6_VOL_DWB), `:=`(PRJ_SP6_VOL_DWB, 0)]

  # Convert projected species volumes to long format
    vol_list2 <- melt(
      vol_list,
      id.vars = c("FEATURE_ID", "PRJ_TOTAL_AGE"),
      measure.vars = c(
        "PRJ_SP1_VOL_DWB",
        "PRJ_SP2_VOL_DWB",
        "PRJ_SP3_VOL_DWB",
        "PRJ_SP4_VOL_DWB",
        "PRJ_SP5_VOL_DWB",
        "PRJ_SP6_VOL_DWB"
      ),
      variable.name = "spcode",
      value.name = "vol"
    )

  # Standardize projected volume species names
  vol_list2[spcode %in% "PRJ_SP1_VOL_DWB", `:=`(spcode, "sp1")]
  vol_list2[spcode %in% "PRJ_SP2_VOL_DWB", `:=`(spcode, "sp2")]
  vol_list2[spcode %in% "PRJ_SP3_VOL_DWB", `:=`(spcode, "sp3")]
  vol_list2[spcode %in% "PRJ_SP4_VOL_DWB", `:=`(spcode, "sp4")]
  vol_list2[spcode %in% "PRJ_SP5_VOL_DWB", `:=`(spcode, "sp5")]
  vol_list2[spcode %in% "PRJ_SP6_VOL_DWB", `:=`(spcode, "sp6")]

  # Merge species and volume tables
    all <- merge(
      sp_list2,
      vol_list2,
      by = c("FEATURE_ID", "spcode"),
      all.x = TRUE
    )

    all2 <- all[
      ,
      .(PRJ_VOL_DWB_conifer = sum(vol, na.rm = TRUE)),
      by = c("FEATURE_ID", "PRJ_TOTAL_AGE")
    ]

  return(all2)
}
```

## Load VDYP Yield Table
```{r load-vdyp}
yld <- readRDS(
  "D:\\data\\projects\\VDYP\\VDYP\\2024\\prov\\VDYP7RunParallel\\prov_lesschanges_nsyt_yieldtable.rds"
)

setDT(yld)
```

## Extract Species Codes
Convert species fields into long format.
```{r species}
d <- unique(
  yld[, .(
    FEATURE_ID,
    SPECIES_1_CODE,
    SPECIES_2_CODE,
    SPECIES_3_CODE,
    SPECIES_4_CODE,
    SPECIES_5_CODE,
    SPECIES_6_CODE
  )]
)

d2 <- melt(
  d,
  id.vars = "FEATURE_ID",
  measure.vars = c(
    "SPECIES_1_CODE",
    "SPECIES_2_CODE",
    "SPECIES_3_CODE",
    "SPECIES_4_CODE",
    "SPECIES_5_CODE",
    "SPECIES_6_CODE"
  ),
  variable.name = "spcode",
  value.name = "sp"
)
```

## Extract Projected Volumes
```{r projected-volumes}
e <- unique(
  yld[, .(
    FEATURE_ID,
    PRJ_TOTAL_AGE,
    PRJ_SP1_VOL_DWB,
    PRJ_SP2_VOL_DWB,
    PRJ_SP3_VOL_DWB,
    PRJ_SP4_VOL_DWB,
    PRJ_SP5_VOL_DWB,
    PRJ_SP6_VOL_DWB
  )]
)

t <- unique(
  yld[, .(
    FEATURE_ID,
    PRJ_TOTAL_AGE,
    PRJ_VOL_DWB
  )]
)
```

## Calculate Conifer and Deciduous Volume
```{r calculate-volumes}
f <- NoDecidVol_calculator(
  sp_list = d2,
  vol_list = e,
  decidsp = decid
)

final <- t %>%
  left_join(f, by = join_by(FEATURE_ID, PRJ_TOTAL_AGE)) %>%
  mutate(
    PRJ_VOL_DWB = round(PRJ_VOL_DWB, 2),
    PRJ_VOL_DWB_conifer = round(PRJ_VOL_DWB_conifer, 2),

    PRJ_VOL_DWB_decid =
      PRJ_VOL_DWB - coalesce(PRJ_VOL_DWB_conifer, 0)
  )
```

## Convert Volume Table to Wide Format

```{r wide-volume}
final_wide <- final %>%
  mutate(
    age_label = paste0("_", PRJ_TOTAL_AGE)
  ) %>%
  pivot_wider(
    id_cols = FEATURE_ID,
    names_from = age_label,
    values_from = c(
      PRJ_VOL_DWB_conifer,
      PRJ_VOL_DWB_decid
    ),
    names_glue =
      "{ifelse(.value == 'PRJ_VOL_DWB_conifer', 'MVcon', 'MVdec')}{age_label}"
  )
```

## Load MSYT Tables

```{r load-msyt}
msyt <- read.csv(
  "D://data//projects//MSYTDelivery//msyt_2025//prov//MSYT_prov_current_input_output.csv"
)

reference <- read.csv(
  "D://data//projects//MSYTDelivery//msyt_2025//prov//MSYT_prov_reference.csv"
)

msyt_fut_in <- read.csv(
  "D://data//projects//MSYTDelivery//msyt_2025//prov//MSYT_prov_future_input.csv"
)
```

## Extend MSYT Curves to Age 250

Ages greater than 120 are populated using age-120 values.

```{r extend-msyt}
new_ages <- seq(130, 250, 10)

msyt_expanded <- msyt

for (age in new_ages) {

  msyt_expanded <- msyt_expanded %>%
    mutate(
      !!paste0("MVcon_", age) := MVcon_120,
      !!paste0("MVdec_", age) := MVdec_120,
      !!paste0("HTcon_", age) := HTcon_120,
      !!paste0("HTdec_", age) := HTdec_120
    )
}
```

## Reorder MSYT Columns

```{r reorder-msyt}
ages <- seq(0, 250, 10)

ordered_cols <- c(
  "feature_id",
  paste0("MVcon_", ages),
  paste0("MVdec_", ages),
  paste0("HTcon_", ages),
  paste0("HTdec_", ages)
)

msyt_expanded <- msyt_expanded %>%
  select(all_of(ordered_cols))
```

## Build MSYT Output

```{r build-msyt}
msyt_part <- reference %>%
  filter(!current_yield %in% c("VDYP", "Excluded")) %>%
  select("feature_id") %>%
  left_join(msyt_expanded, by = "feature_id") %>%
  mutate(source = "MSYT")
```

## Load Height Ratios

```{r load-height-ratios}
ht_ratio <- read.csv(
  "D://data//projects//VDYP//VDYP//2024//prov//bec_height_ratios.csv"
)
```

## Calculate VDYP Heights
Calculate Conifer and Deciduous Heights based on the ratio table

```{r vdyp-heights}
vdyp_ht <- unique(
  yld[, .(
    FEATURE_ID,
    PRJ_TOTAL_AGE,
    PRJ_DOM_HT,
    SPECIES_1_CODE
  )]
)

vdyp_ht <- vdyp_ht %>%
  left_join(
    reference %>%
      select(feature_id, bec_zone, bec_subzone) %>%
      mutate(
        bec_label = paste0(bec_zone, "_", bec_subzone)
      ),
    by = c("FEATURE_ID" = "feature_id")
  )

vdyp_ht <- vdyp_ht %>%
  left_join(ht_ratio, by = "bec_label")

vdyp_ht <- vdyp_ht %>%
  mutate(
    is_decid = SPECIES_1_CODE %in% decid,

    HTcon = case_when(
      is_decid ~ PRJ_DOM_HT * con_dec_ratio,
      TRUE ~ PRJ_DOM_HT
    ),

    HTdec = case_when(
      is_decid ~ PRJ_DOM_HT,
      TRUE ~ PRJ_DOM_HT * dec_con_ratio
    )
  )

vdyp_ht_wide <- vdyp_ht %>%
  mutate(
    age_label = paste0("_", PRJ_TOTAL_AGE)
  ) %>%
  pivot_wider(
    id_cols = FEATURE_ID,
    names_from = age_label,
    values_from = c(HTcon, HTdec),
    names_glue = "{.value}{age_label}"
  )
```

## Build VDYP Output

```{r build-vdyp}
vdyp_part <- reference %>%
  filter(current_yield %in% c("VDYP", "Excluded")) %>%
  select("feature_id") %>%
  left_join(
    final_wide,
    by = c("feature_id" = "FEATURE_ID")
  ) %>%
  left_join(
    vdyp_ht_wide,
    by = c("feature_id" = "FEATURE_ID")
  ) %>%
  mutate(source = "NSYT")

final_yld <- bind_rows(
  msyt_part,
  vdyp_part
)
```

## Export Final Yield Table

```{r export}
write.csv(
  final_yld,
  "D:\\data\\projects\\VDYP\\VDYP\\2024\\prov\\yld_combined.csv",
  row.names = FALSE
)
```

## QA/QC Checks

Identify records where all merchantable volume columns equal 0.

```{r qaqc}
mv_cols <- grep(
  "^MV(con|dec)_",
  names(final_yld),
  value = TRUE
)

zero_features <- final_yld %>%
  mutate(
    total_mv = rowSums(
      across(all_of(mv_cols)),
      na.rm = TRUE
    )
  ) %>%
  filter(total_mv == 0) %>%
  select(feature_id, source)
```

## Review NSYT Zero-Volume Records

```{r nsyt-zero}
nsyt_all_zero <- zero_features %>%
  filter(source == "NSYT") %>%
  left_join(reference, by = "feature_id") %>%
  left_join(t, by = c("feature_id" = "FEATURE_ID"))

nsyt_all_zero
```

## Review MSYT Zero-Volume Records

```{r msyt-zero}
msyt_all_zero <- zero_features %>%
  filter(source == "MSYT") %>%
  left_join(reference, by = "feature_id") %>%
  left_join(msyt, by = "feature_id")

msyt_all_zero
```
