# Comparison with VRI R1

The comparison process was designed to assess the consistency between the Beta version of the 2026 Provincial NSYT dataset and the currently published VRI R1. The workflow begins by loading the provincial NSYT yield table and joining it to the R1 inventory database using a common feature_id. Species, age, and volume information from both sources are combined to create a common analytical framework. To facilitate a direct comparison, stand ages are standardized, volume estimates are derived from the R1 inventory, and VDYP yield projections are interpolated to match the specific ages represented in the inventory. This allows both datasets to be evaluated using comparable volume estimates for the same forest features.

A difference metric is calculated to quantify the variation between the inventory-derived volume (volume from VRI R1 layer) and the NSYT volume estimate (interpolated NSYT volume). The resulting dataset provides a feature-by-feature comparison of the two yield systems and supports statistical evaluation of differences across the provincial land base. This process serves as a validation and quality assurance exercise, helping identify areas where modelling assumptions, age treatments, interpolation methods, or source inventory data may contribute to differences between the existing R1 inventory volumes and the proposed provincial NSYT yield dataset.

The following contrasts the Beta version of the 2026 NSYT dataset with the currently published production year 2025 (projection year Dec 2024) veg_comp_lyr_r1_poly_2024 (R1). The R1 contains **6872386** features, 90% of which (**6209645**) has a species identified and  (spec_cd_1) and 89% (**6140967**) remain after initial cleaning.

```{r set-up, message=FALSE,warning=FALSE,eval=FALSE}
library(DBI)
library(RPostgreSQL)
library(tidyverse)
library(terra)
library(R.utils)
library(tidytext)
library(arsenal)
library(dlookr)
library(gdalUtilities)
library(glue)
library(janitor)
library(knitr)
library(kableExtra)
library(tidytext)
library(fs)
library(data.table)
library(glue)
library(janitor)
library(dplyr)

rm(list = ls())


options(scipen = 999)
set.seed(123)
# source("log_files.R")


knitr::opts_chunk$set(
  echo = TRUE,              # Show code in the output
  fig.width = 4,            # Set the width of the figures (in inches)
  fig.height = 4,           # Set the height of the figures (in inches)
  collapse = TRUE,          # Collapsing code and output
  # out.width = '250px',      # Standardize output image width
  # out.height = '250px',     # Standardize output image height
  message = FALSE,          # Hide messages
  warning = FALSE           # Hide warnings
)


```
##  Get data

The following snippet loads 2026 provincial NSYT from an RDS file containing **6140967** features, and joins it with the data pulled from the VRI vegetation composition R1 layer based on feature_id. Various new columns are created, including volumes from VRI R1 layer(vri_vol), capped age (vri_age250), and interpolated volumes (vdyp_vol) based on age ranges. 

```{r get_data, message=FALSE,warning=FALSE,eval=FALSE}

schema <- 'msyt_2025'
opt <- paste0("-c search_path=",schema)
user_name <- 'results'
database <- 'msyt'
db<- dbConnect('PostgreSQL',dbname=database,user=user_name,options = opt)

prov<-readRDS("./data/rds/prov_nsyt2026.rds")


sql<-glue_sql("select
              feature_id,
              proj_age_1 as age,
              species_cd_1 as leading_species,
              species_cd_2 as second_species,
              live_stand_volume_125 as r1_vol125,
              live_stand_volume_175 as r1_vol175,
              dead_stand_volume_125 as r1_dvol125
              from veg_comp_fdw_full",.con = db)

r1<-dbGetQuery(db,sql)

##############################################################################
# # 
prov_df<-inner_join(prov_df,r1,join_by("feature_id"))

prov_df2 <- prov_df %>%
  mutate(
    vri_vol = case_when(
      grepl("^P", leading_species) ~ r1_vol125,
      grepl("^A", leading_species) ~ r1_vol125,
      grepl("^E", leading_species) ~ r1_vol125,
      grepl("^D", leading_species) ~ r1_vol125,
      grepl("^M", leading_species) ~ r1_vol125,
      grepl("^W", leading_species) ~ r1_vol125,
      TRUE ~ r1_vol175  # Otherwise, assign r1_vol175
    ),
    across(where(is.numeric), ~replace_na(., 0)),
    vri_age250 = pmin(age, 250),
    age_low  = floor(vri_age250/10) * 10,
    age_high = ceiling(vri_age250/10) * 10,
    p_low    = (age_high - vri_age250) / 10
  ) %>%
  group_by(feature_id) %>%
  mutate(
    l_vol = prj_vol_dwb[match(age_low,  prj_total_age)],
    u_vol = prj_vol_dwb[match(age_high, prj_total_age)],
    vdyp_vol = (p_low * l_vol) + ((1 - p_low) * u_vol)
  ) %>%
  select(feature_id, vri_vol,l_vol, u_vol, vdyp_vol)%>%
  slice(1)%>%
  ungroup()

nsyt<-inner_join(prov_df,prov_df2,join_by("feature_id"))

```

##  Closest Match

The closest match volume calculation works by identifying the record with the smallest absolute difference between two volume estimates for each feature.  

- Filtering for Valid Volumes: The script first filters out any rows where either vdyp_vol or prj_vol_dwb (the decadal VDYP volume) is NA, ensuring that only valid rows with both volume estimates are considered for comparison.

- Grouping by feature_id: The data is grouped by feature_id, meaning the script performs the closest match calculation separately for each feature (or record).

- Identifying the Closest Match: For each group (feature_id), the script calculates the absolute difference between vdyp_vol and prj_vol_dwb using the expression abs(prj_vol_dwb - vdyp_vol). The row with the smallest difference is considered the "closest match" for that feature. This is achieved using the slice_min function, which selects the row where the absolute difference is the smallest (n = 1), ensuring that ties (rows with the same smallest difference) are excluded (with_ties = FALSE).

- Calculating the Volume Difference: Once the closest match is found, a new column vol_dif is created, which represents the difference between the VDYP derived lineral interpolation and the R1 volume: vol_dif = vri_vol - vdyp_vol.The output is then saved as a RDS file.

The resulting NSYT closest data frame is loaded to disk and its vol_dif field is then assessed statistically using the dlookr::describe function. 

```{r closest, eval=FALSE}
nsyt_closest <- nsyt %>%
  filter(!is.na(vdyp_vol), !is.na(prj_vol_dwb)) %>%
  group_by(feature_id) %>%
  slice_min(abs(prj_vol_dwb - vdyp_vol),
            n = 1,
            with_ties = FALSE) %>%
  ungroup()%>%
  mutate(vol_dif = vri_vol - vdyp_vol)
```

```{r tab-closest_match,message=FALSE,warning=FALSE,eval=TRUE,echo=FALSE}
library(dlookr)
library(dplyr)
library(knitr)
library(kableExtra)

prov<-readRDS("../nsyt_closest.rds")
# pretty_table(describe(prov))
kable(describe(prov),
      caption =  "Volume Comparison Table",
      format = "html",
      booktabs = TRUE) %>%
 kable_styling(full_width = TRUE,
               font_size = 10) %>%
  scroll_box(width = "100%")

```

Of the **6140967** features assessed the mean difference in volume between the VDYP dataset and the R1 is **-1.8m3 (sd: 11.13)**, the median difference is **-0.015m3** , **986984** (16%) of the features have volume difference > +/- 5m3, **454585** (8%) features have volume difference >  +/- 10m3 and **170037** (2.7%) features have volume difference > +/- 20m3.

The are a number of sources that drive divergence including approaches to interpolation (most common source for volume differences < 5m3), age cutoffs and layer/project table input error in both the R1 and vdyp_vol dataset. The provincial VDYP dataset has an age cutoff of 250 years effecting **244710** (25%) of the **986984**  of the features have volume difference > +/- 5m3. 85% of the features > 250 years of age have vdyp volumes > the R1 volume suggesting the much older stands in the R1 are decrementing volume, driving the divergence. There are **15270** features that have ages > 250 and vdyp_vols == 0. 

Of the features < 250 years of age **742274** (12%) of the features have volume difference > +/- 5m3, **285508** (4.6%) features have volume difference >  +/- 10m3 and **70880** (1%) features have volume difference > +/- 20m3. There **9935** features where the vri_vol == 0 and the vdyp_vol > 0 and **365** features where the vdyp_vol  == 0.

**112262** features < 250 years have R1 yields more reflective of the layer table est_age_1 and 2 fields than the projected age (proj_age_1) suggesting an error has occured in the 2025 update.  

